Revisions

### 2024-08-27 Manuscript/manuscript.qmd
fix minor typo
add discussion of model comparison between projects; and stimuli vs. response issue. Fix $c$ formatting.

To directly control for similarity in Experiment 2, we developed a model-based measure of the similarity between training throws and testing conditions. This similarity measure was a significant predictor of testing performance, e.g., participants whose training throws were more similar to throws that resulted in target hits from the testing positions, tended to perform better during the testing phase. Importantly, the similarity measure did not explain away the group-level benefits of varied training, which remained significant in our linear model predicting testing performance after similarity was added to the model. However, previous research has suggested that participants may differ in their level of generalization as a function of prior experience, and that such differences in generalization gradients can be captured by fitting the generalization parameter of an instance-based model separately to each group [@hahnEffectsCategoryDiversity2005; @lambertsFlexibleTuningSimilarity1994]. Relatedly, the influential Bayesian generalization model developed by @tenenbaumGeneralizationSimilarityBayesian2001a predicts that the breadth of generalization will increase when a rational agent encounters a wider variety of examples. Following these leads, we assume that in addition to learning the task itself, participants are also adjusting how generalizable their experience should be. Varied versus constant participants may be expected to learn to generalize their experience to different degrees. To accommodate this difference, the generalization parameter of the instance-based model (in the present case, the c parameter) can be allowed to vary between the two groups to reflect the tendency of learners to adaptively tune the extent of their generalization. One specific hypothesis is that people adaptively set a value of c to fit the variability of their training experience [@nosofskyExemplarbasedAccountsMultiplesystem2000; @sakamotoTrackingVariabilityLearning2006]. If one's training experience is relatively variable, as with the variable training condition, then one might infer that future test situations will also be variable, in which case a low value of c will allow better generalization because generalization will drop off slowly with training-to-testing distance. Conversely, if one's training experience has little variability, as found in the constant training conditions, then one might adopt a high value of c so that generalization falls off rapidly away from the trained positions.

@@

Linear models 1 and 3 both show that similarity is a significant predictor of testing performance (p\<.01). Of greater interest is the difference between linear model 2, in which similarity is computed from a single c value fit from all participants (Similarity1c), with linear model 4, which fits the c parameter separately between groups (Similarity2c). In linear model 2, the effect of training group remains significant when controlling for Similarity1c (p\<.01), with the varied group still performing significantly better. However, in linear model 4 the addition of the Similarity2c predictor results in the effect of training group becoming nonsignificant (p=.40), suggesting that the effect of varied vs. constant training is accounted for by the Similarity2c predictor. Next, to further establish a difference between the models, we performed nested model comparisons using ANOVA, to see if the addition of the training group parameter led to a significant improvement in model performance. In the first comparison, ANOVA(Linear Model 1, Linear Model 2), the addition of the training group predictor significantly improved the performance of the model (F=22.07, p\<.01). However, in the second model comparison, ANOVA (Linear model 3, Linear Model 4) found no improvement in model performance with the addition of the training group predictor (F=1.61, p=.20).

Finally, we sought to confirm that similarity values generated from the adjusted Similarity2c model had more predictive power than those generated from the original Similarity1c model. Using the BIC function in R, we compared BIC values between linear model 1 (BIC=14604.00) and linear model 3 (BIC = 14587.64). The lower BIC value of model 3 suggests a modest advantage for predicting performance using a similarity measure computed with two c values over similarity computed with a single c value. When fit with separate c values, the best fitting c parameters for the model consistently optimized such that the c value for the varied group (c=.00008) was smaller in magnitude than the c value for the constant group (c= .00011). Recall that similarity decreases as a Gaussian function of distance (equation 1 above), and a smaller value of c will result in a more gradual drop-off in similarity as the distance between training throws and testing solutions increases.

@@

Our results thus suggest that the benefits of variation cannot be explained by the varied-trained participants simply covering a broader range of the task space. Rather, the modeling suggests that varied participants also learn to adaptively tune their generalization function such that throwing locations generalize more broadly to one another than they do in the constant condition. A learning system could end up adopting a higher c value in the constant than variable training conditions by monitoring the trial-by-trial variability of the training items. The c parameter would be adapted downwards when adjacent training items are dissimilar to each other and adapted upwards when adjacent training items are the same. In this fashion, contextually appropriate c values could be empirically learned. This learning procedure would capture the insight that if a situation has a high amount variability, then the learner should be predisposed toward thinking that subsequent test items will also show considerable variability, in which case generalization gradients should be broad, as is achieved by low values for c. @sakamotoTrackingVariabilityLearning2006 implemented a similar learning mechanism for updating the generalization paramater in an exemplar-based model (although in their model, a separate generalization parameter is assigned to each exemplar). In their experiment, participants were trained on a high variability and a low variability category, and the dynamically updated generalization parameter was necessary to account for broader generalization observed around the high variability category when participants were tested with an ambiguous intermediary item. In a subsequent work [@sakamotoPuttingPsychologyBack2008], the same authors showed that a similar learning mechanism could account for the pattern wherein participants generalize more broadly around a category when the average distance between the category exemplars is larger (however the only model tested in this work was a prototype model).

@@

```{r}
#| label: fig-htw-post-dist
#| fig-cap: "Posterior Distributions of $c$ and $lr$ parameters. Points represent median values, thicker intervals represent 66% credible intervals and thin intervals represent 95% credible intervals around the median. Note that the y-axes of the plots for the c parameter are scaled logarithmically."
#| fig-height: 6
#| fig-width: 10

@@

Although theoretically motivated, the IGAS model of Project 1 is best categorized as a descriptive measurement-model. Sufficient to account for group differences, but lacking the machinery necessary to provide a full process-level account of how the empirical quantities of interest are generated. In contrast, Project 2 (HTW) implemented a more robust computational modeling approach, implementing and comparing full process models (ALM & EXAM), capable of generating predictions for both the learning and testing stages of the experiment. ALM and EXAM have been used as models of function learning, cue judgement, and forecasting behavior in numerous studies over the past 25 years [@deloshExtrapolationSineQua1997; @mcdanielPredictingTransferPerformance2009; @brownUnderestimationLinearFunction2017; @kwantesItemOrderMatters2012; @kelleyComparisonModelsLearning2008; @kaneApplicationsBiasVariance2020; @vonhelversenLearningMultiplecueJudgment2010]. The present work presents the first application of these models to to the study of training variability in a visuomotor function learning task. We fit both models to individual participant data, using a form of simulation-based Bayesian parameter estimation that allowed us to generate and compare the full posterior predictive distributions of each model. EXAM provided the best overall account of the testing data, and the advantage of EXAM over ALM was significantly greater for the constant group. Notably, EXAM captured the constant groups' ability to extrapolate linearly to novel velocity bands, despite receiving training from only a single input-output pair. This finding suggests that EXAM's linear extrapolation mechanism, combined with the assumption of prior knowledge about the origin point (0, 0), was sufficient to account for the constant groups' accurate extrapolation performance. Such findings may offer a preliminary suggestion that experience with a more variable set of training examples  may be detrimental to performance in simple extrapolation tasks.

@@

The HTT and HTW tasks differ across numerous dimensions that may be relevant to the opposing patterns observed in the two projects (see @tbl-task-diff provides for a detailed comparison of the two tasks).

@@

>>>>

To directly control for similarity in Experiment 2, we developed a model-based measure of the similarity between training throws and testing conditions. This similarity measure was a significant predictor of testing performance, e.g., participants whose training throws were more similar to throws that resulted in target hits from the testing positions, tended to perform better during the testing phase. Importantly, the similarity measure did not explain away the group-level benefits of varied training, which remained significant in our linear model predicting testing performance after similarity was added to the model. However, previous research has suggested that participants may differ in their level of generalization as a function of prior experience, and that such differences in generalization gradients can be captured by fitting the generalization parameter of an instance-based model separately to each group [@hahnEffectsCategoryDiversity2005; @lambertsFlexibleTuningSimilarity1994]. Relatedly, the influential Bayesian generalization model developed by @tenenbaumGeneralizationSimilarityBayesian2001a predicts that the breadth of generalization will increase when a rational agent encounters a wider variety of examples. Following these leads, we assume that in addition to learning the task itself, participants are also adjusting how generalizable their experience should be. Varied versus constant participants may be expected to learn to generalize their experience to different degrees. To accommodate this difference, the generalization parameter of the instance-based model (in the present case, the $c$ parameter) can be allowed to vary between the two groups to reflect the tendency of learners to adaptively tune the extent of their generalization. One specific hypothesis is that people adaptively set a value of c to fit the variability of their training experience [@nosofskyExemplarbasedAccountsMultiplesystem2000; @sakamotoTrackingVariabilityLearning2006]. If one's training experience is relatively variable, as with the variable training condition, then one might infer that future test situations will also be variable, in which case a low value of c will allow better generalization because generalization will drop off slowly with training-to-testing distance. Conversely, if one's training experience has little variability, as found in the constant training conditions, then one might adopt a high value of c so that generalization falls off rapidly away from the trained positions.

@@

Linear models 1 and 3 both show that similarity is a significant predictor of testing performance (p\<.01). Of greater interest is the difference between linear model 2, in which similarity is computed from a single c value fit from all participants (Similarity1c), with linear model 4, which fits the $c$ parameter separately between groups (Similarity2c). In linear model 2, the effect of training group remains significant when controlling for Similarity1c (p\<.01), with the varied group still performing significantly better. However, in linear model 4 the addition of the Similarity2c predictor results in the effect of training group becoming nonsignificant (p=.40), suggesting that the effect of varied vs. constant training is accounted for by the Similarity2c predictor. Next, to further establish a difference between the models, we performed nested model comparisons using ANOVA, to see if the addition of the training group parameter led to a significant improvement in model performance. In the first comparison, ANOVA(Linear Model 1, Linear Model 2), the addition of the training group predictor significantly improved the performance of the model (F=22.07, p\<.01). However, in the second model comparison, ANOVA (Linear model 3, Linear Model 4) found no improvement in model performance with the addition of the training group predictor (F=1.61, p=.20).

Finally, we sought to confirm that similarity values generated from the adjusted Similarity2c model had more predictive power than those generated from the original Similarity1c model. Using the BIC function in R, we compared BIC values between linear model 1 (BIC=14604.00) and linear model 3 (BIC = 14587.64). The lower BIC value of model 3 suggests a modest advantage for predicting performance using a similarity measure computed with two c values over similarity computed with a single c value. When fit with separate c values, the best fitting $c$ parameters for the model consistently optimized such that the c value for the varied group (c=.00008) was smaller in magnitude than the c value for the constant group (c= .00011). Recall that similarity decreases as a Gaussian function of distance (equation 1 above), and a smaller value of c will result in a more gradual drop-off in similarity as the distance between training throws and testing solutions increases.

@@

Our results thus suggest that the benefits of variation cannot be explained by the varied-trained participants simply covering a broader range of the task space. Rather, the modeling suggests that varied participants also learn to adaptively tune their generalization function such that throwing locations generalize more broadly to one another than they do in the constant condition. A learning system could end up adopting a higher c value in the constant than variable training conditions by monitoring the trial-by-trial variability of the training items. The $c$ parameter would be adapted downwards when adjacent training items are dissimilar to each other and adapted upwards when adjacent training items are the same. In this fashion, contextually appropriate c values could be empirically learned. This learning procedure would capture the insight that if a situation has a high amount variability, then the learner should be predisposed toward thinking that subsequent test items will also show considerable variability, in which case generalization gradients should be broad, as is achieved by low values for c. @sakamotoTrackingVariabilityLearning2006 implemented a similar learning mechanism for updating the generalization paramater in an exemplar-based model (although in their model, a separate generalization parameter is assigned to each exemplar). In their experiment, participants were trained on a high variability and a low variability category, and the dynamically updated generalization parameter was necessary to account for broader generalization observed around the high variability category when participants were tested with an ambiguous intermediary item. In a subsequent work [@sakamotoPuttingPsychologyBack2008], the same authors showed that a similar learning mechanism could account for the pattern wherein participants generalize more broadly around a category when the average distance between the category exemplars is larger (however the only model tested in this work was a prototype model).

@@

```{r}
#| label: fig-htw-post-dist
#| fig-cap: "Posterior Distributions of $c$ and $lr$ parameters. Points represent median values, thicker intervals represent 66% credible intervals and thin intervals represent 95% credible intervals around the median. Note that the y-axes of the plots for the $c$ parameter are scaled logarithmically."
#| fig-height: 6
#| fig-width: 10

@@

Although theoretically motivated, the IGAS model of Project 1 is best categorized as a descriptive measurement-model. Sufficient to account for group differences, but lacking the machinery necessary to provide a full process-level account of how the empirical quantities of interest are generated. In contrast, Project 2 (HTW) implemented a more robust computational modeling approach, implementing and comparing full process models (ALM & EXAM), capable of generating predictions for both the learning and testing stages of the experiment. ALM and EXAM have been used as models of function learning, cue judgement, and forecasting behavior in numerous studies over the past 25 years [@deloshExtrapolationSineQua1997; @mcdanielPredictingTransferPerformance2009; @brownUnderestimationLinearFunction2017; @kwantesItemOrderMatters2012; @kelleyComparisonModelsLearning2008; @kaneApplicationsBiasVariance2020; @vonhelversenLearningMultiplecueJudgment2010]. The present work presents the first application of these models to the study of training variability in a visuomotor function learning task. We fit both models to individual participant data, using a form of simulation-based Bayesian parameter estimation that allowed us to generate and compare the full posterior predictive distributions of each model. EXAM provided the best overall account of the testing data, and the advantage of EXAM over ALM was significantly greater for the constant group. Notably, EXAM captured the constant groups' ability to extrapolate linearly to novel velocity bands, despite receiving training from only a single input-output pair. This finding suggests that EXAM's linear extrapolation mechanism, combined with the assumption of prior knowledge about the origin point (0, 0), was sufficient to account for the constant groups' accurate extrapolation performance. Such findings may offer a preliminary suggestion that experience with a more variable set of training examples  may be detrimental to performance in simple extrapolation tasks.

@@

The HTT and HTW tasks differ across numerous dimensions that may be relevant to the opposing patterns observed in the two projects (see @tbl-task-diff for a detailed comparison of the two tasks).

@@

It is important to note that while both projects utilize computational models, direct comparisons are complicated by the distinct purposes and structures of the models employed in each project. The IGAS model of Project 1 serves as a descriptive measurement model, capturing the similarity between training throws and testing conditions. In contrast, the ALM and EXAM models of Project 2 are full process models, capable of generating exact predictions for both learning and testing stages. The difference is also reflected in the interpretion of the generalization parameter ($c$) across the models of the two projects. In IGAS, $c$ moderates the similarity between executed throws and subsequent testing solutions, while in ALM and EXAM, $c$ governs the extent to which the perceived stimuli activate the input layer nodes. Despite these differences, insights from ALM/EXAM, particularly the role of zero-point knowledge, may offer potential explanations for the contrasting empirical results. Particularly, EXAM's reliance on zero-point knowledge in the simpler HTW task may explain why constant training was more effective in Project 2, while the lack of a clear zero-point reference in the more complex HTT task of Project 1 may have increased the value of varied training. This suggests that the benefits of variability depend critically on how task structure interacts with prior knowledge and the learner's capacity to leverage such knowledge for generalization.

Future work could explore extending ALM and EXAM, which have traditionally been applied to one-dimensional function learning tasks, to more complex motor tasks such as HTT. The neural network structure of ALM could be adapted to handle 2D input by utilizing a 2D grid of input nodes, allowing the model to learn mappings between 2D throwing velocities and desired outcomes. This would allow the model to process the more complex spatial information inherent in tasks like HTT. Furthermore, the output layers of ALM/EXAM could be expanded to express more complex motor outputs in addition to velocity, such as the locations of grabbing and releasing the projectile or other parameters defining the unique trajectories produced. In addition to allowing the models to be applied to more complex tasks, these modifications could enable researchers to investigate how perceptual similarity (i.e., the similarity of stimuli) and motoric similarity (i.e., the similarity of behavioral actions) may separately and jointly influence learning and generalization.

## Conclusion

### 2024-08-27 Sections/Discussion.qmd
fix minor typo
add discussion of model comparison between projects; and stimuli vs. response issue. Fix $c$ formatting.

### Conclusion

>>>>

It is important to note that while both projects utilize computational models, direct comparisons are complicated by the distinct purposes and structures of the models employed in each project. The IGAS model of Project 1 serves as a descriptive measurement model, capturing the similarity between training throws and testing conditions. In contrast, the ALM and EXAM models of Project 2 are full process models, capable of generating exact predictions for both learning and testing stages. The difference is also reflected in the interpretion of the generalization parameter ($c$) across the models of the two projects. In IGAS, $c$ moderates the similarity between executed throws and subsequent testing solutions, while in ALM and EXAM, $c$ governs the extent to which the perceived stimuli activate the input layer nodes. Despite these differences, insights from ALM/EXAM, particularly the role of zero-point knowledge, may offer potential explanations for the contrasting empirical results. Particularly, EXAM's reliance on zero-point knowledge in the simpler HTW task may explain why constant training was more effective in Project 2, while the lack of a clear zero-point reference in the more complex HTT task of Project 1 may have increased the value of varied training. This suggests that the benefits of variability depend critically on how task structure interacts with prior knowledge and the learner's capacity to leverage such knowledge for generalization.

Future work could explore extending ALM and EXAM, which have traditionally been applied to one-dimensional function learning tasks, to more complex motor tasks such as HTT. The neural network structure of ALM could be adapted to handle 2D input by utilizing a 2D grid of input nodes, allowing the model to learn mappings between 2D throwing velocities and desired outcomes. This would allow the model to process the more complex spatial information inherent in tasks like HTT. Furthermore, the output layers of ALM/EXAM could be expanded to express more complex motor outputs in addition to velocity, such as the locations of grabbing and releasing the projectile or other parameters defining the unique trajectories produced. In addition to allowing the models to be applied to more complex tasks, these modifications could enable researchers to investigate how perceptual similarity (i.e., the similarity of stimuli) and motoric similarity (i.e., the similarity of behavioral actions) may separately and jointly influence learning and generalization.

## Conclusion

### 2024-08-27 Sections/IGAS.qmd
add discussion of model comparison between projects; and stimuli vs. response issue. Fix $c$ formatting.

Linear models 1 and 3 both show that similarity is a significant predictor of testing performance (p\<.01). Of greater interest is the difference between linear model 2, in which similarity is computed from a single c value fit from all participants (Similarity1c), with linear model 4, which fits the c parameter separately between groups (Similarity2c). In linear model 2, the effect of training group remains significant when controlling for Similarity1c (p\<.01), with the varied group still performing significantly better. However, in linear model 4 the addition of the Similarity2c predictor results in the effect of training group becoming nonsignificant (p=.40), suggesting that the effect of varied vs. constant training is accounted for by the Similarity2c predictor. Next, to further establish a difference between the models, we performed nested model comparisons using ANOVA, to see if the addition of the training group parameter led to a significant improvement in model performance. In the first comparison, ANOVA(Linear Model 1, Linear Model 2), the addition of the training group predictor significantly improved the performance of the model (F=22.07, p\<.01). However, in the second model comparison, ANOVA (Linear model 3, Linear Model 4) found no improvement in model performance with the addition of the training group predictor (F=1.61, p=.20).

Finally, we sought to confirm that similarity values generated from the adjusted Similarity2c model had more predictive power than those generated from the original Similarity1c model. Using the BIC function in R, we compared BIC values between linear model 1 (BIC=14604.00) and linear model 3 (BIC = 14587.64). The lower BIC value of model 3 suggests a modest advantage for predicting performance using a similarity measure computed with two c values over similarity computed with a single c value. When fit with separate c values, the best fitting c parameters for the model consistently optimized such that the c value for the varied group (c=.00008) was smaller in magnitude than the c value for the constant group(c= .00011). Recall that similarity decreases as a Gaussian function of distance (equation 1 above), and a smaller value of c will result in a more gradual drop-off in similarity as the distance between training throws and testing solutions increases.

>>>>

Linear models 1 and 3 both show that similarity is a significant predictor of testing performance (p\<.01). Of greater interest is the difference between linear model 2, in which similarity is computed from a single c value fit from all participants (Similarity1c), with linear model 4, which fits the $c$ parameter separately between groups (Similarity2c). In linear model 2, the effect of training group remains significant when controlling for Similarity1c (p\<.01), with the varied group still performing significantly better. However, in linear model 4 the addition of the Similarity2c predictor results in the effect of training group becoming nonsignificant (p=.40), suggesting that the effect of varied vs. constant training is accounted for by the Similarity2c predictor. Next, to further establish a difference between the models, we performed nested model comparisons using ANOVA, to see if the addition of the training group parameter led to a significant improvement in model performance. In the first comparison, ANOVA(Linear Model 1, Linear Model 2), the addition of the training group predictor significantly improved the performance of the model (F=22.07, p\<.01). However, in the second model comparison, ANOVA (Linear model 3, Linear Model 4) found no improvement in model performance with the addition of the training group predictor (F=1.61, p=.20).

Finally, we sought to confirm that similarity values generated from the adjusted Similarity2c model had more predictive power than those generated from the original Similarity1c model. Using the BIC function in R, we compared BIC values between linear model 1 (BIC=14604.00) and linear model 3 (BIC = 14587.64). The lower BIC value of model 3 suggests a modest advantage for predicting performance using a similarity measure computed with two c values over similarity computed with a single c value. When fit with separate c values, the best fitting $c$ parameters for the model consistently optimized such that the c value for the varied group (c=.00008) was smaller in magnitude than the c value for the constant group(c= .00011). Recall that similarity decreases as a Gaussian function of distance (equation 1 above), and a smaller value of c will result in a more gradual drop-off in similarity as the distance between training throws and testing solutions increases.

### 2024-08-27 revisions.qmd
add discussion of model comparison between projects; and stimuli vs. response issue. Fix $c$ formatting.

## Project Comparison

>>>>

$c$ vs. *c*

## Project Model Comparison/Discussion

It is important to note that while both projects utilize computational models, direct comparisons are complicated by the distinct purposes and structures of the models employed in each project. The IGAS model of Project 1 serves as a descriptive measurement model, capturing the similarity between training throws and testing conditions. In contrast, the ALM and EXAM models of Project 2 are full process models, capable of generating exact predictions for both learning and testing stages. The difference is also reflected in the interpretion of the generalization parameter ($c$) across the models of the two projects. In IGAS, $c$ moderates the similarity between executed throws and subsequent testing solutions, while in ALM and EXAM, $c$ governs the extent to which the perceived stimuli activate the input layer nodes. Despite these differences, insights from ALM/EXAM, particularly the role of zero-point knowledge, may offer potential explanations for the contrasting empirical results. Particularly, EXAM's reliance on zero-point knowledge in the simpler HTW task may explain why constant training was more effective in Project 2, while the lack of a clear zero-point reference in the more complex HTT task of Project 1 may have increased the value of varied training. This suggests that the benefits of variability depend critically on how task structure interacts with prior knowledge and the learner's capacity to leverage such knowledge for generalization.

Future work could explore extending ALM and EXAM, which have traditionally been applied to one-dimensional function learning tasks, to more complex motor tasks such as HTT. The neural network structure of ALM could be adapted to handle 2D input by utilizing a 2D grid of input nodes, allowing the model to learn mappings between 2D throwing velocities and desired outcomes. This would allow the model to process the more complex spatial information inherent in tasks like HTT. Furthermore, the output layers of ALM/EXAM could be expanded to express more complex motor outputs in addition to velocity, such as the locations of grabbing and releasing the projectile or other parameters defining the unique trajectories produced. In addition to allowing the models to be applied to more complex tasks, these modifications could enable researchers to investigate how perceptual similarity (i.e., the similarity of stimuli) and motoric similarity (i.e., the similarity of behavioral actions) may separately and jointly influence learning and generalization.
